A Linear Solution of the Linearity Challenge

### 2025-12-28 paper/related.tex
revision

We have compared our formalisation with others that have appeared in the
literature and for which we were able to retrieve the source code. We have
looked at those formalisations that include the following features:
\begin{itemize}
\item reduction/evaluation semantics for a calculus with support for binary
    sessions, either natively or through their encoding in terms of linear
    channels~\cite{KobayashiPierceTurner99,Kobayashi02b};
\item a type system along with the usual typing preservation results;
\item optionally, a proof of global progress/deadlock freedom.
\end{itemize}

@@

\citet{ZackonEtAl25} describe yet another approach for managing typing contexts
whereby channels are associated not just with a type but also with a \emph{tag}
(or \emph{multiplicity}) from a given resource algebra. The tag summarises the
number of allowed usages of a particular channel, including the possibility that
the channel is not available. This approach streamlines context splitting since
contexts can be treated in an essentially structural way, except for tags which
are conveninently combined using operations from the resource algebra.

@@

\Cref{tab:comparison} shows an overview of the aforementioned formalisations
(sorted by publication date) including our own. The first five columns identify
the calculus being formalised. We provide its reference paper, the language in
which it is formalised and an acronym that gives an idea of the flavour of the
calculus. We also specify whether the calculus features \emph{cuts} (that is,
the combination of restriction and parallel composition that corresponds with
the cut of linear logic) and continuations.
%
\CP~\cite{Wadler14,GayVasconcelos25} and \GV~\cite{GayVasconcelos10,Wadler14}
are well-known acronyms in the literature on session types. \SessionPI refers to
(variants of) the session-based $\pi$-calculus presented by
\citet{HondaVasconcelosKubo98} while \LinearPI refers to (variants of) the
linear $\pi$-calculus~\cite{KobayashiPierceTurner99}. Finally, \SCP is the
structural version of \CP introduced by \citet{SanoKavanaghPientka23} and
\Calculus is our calculus.
%
We emphasize that the actual calculus being formalised usually differs from
(typically, is a strict subset of) the one identified by the acronym and that
the same acronym may refer to different versions of the same calculus. In
particular, \GV in a logical setting is described by \citet{Wadler14} but its
first (non-logical) version is due to \citet{GayVasconcelos10}.

@@

The three middle columns of \Cref{tab:comparison} report the relevant
qualitative aspects of the formalisations, namely the management of typing
contexts, the representation of channels and whether the formalisation is
intrinsically or extrinsically typed.

>>>>

We have compared our formalisation with others of typed calculi that support
binary sessions either natively or through their encoding using
continuations~\cite{KobayashiPierceTurner99,Kobayashi02b}.

@@

\citet{ZackonEtAl25} describe a typing context management technique where
channels are associated not just with a type but also with an element -- called
\emph{tag} -- of a given resource algebra that summarises the number of allowed
usages of a particular channel, including the possibility that the channel is
not available. This approach streamlines context splitting since contexts can be
treated in an essentially structural way, except for tags which are
conveninently combined using operations from the resource algebra.

@@

\Cref{tab:comparison} shows an overview of the aforementioned formalisations
(sorted by publication date) including our own. The first five columns identify
the calculus being formalised. We provide its reference paper, the prover in
which it is formalised and an acronym that gives an idea of the flavour of the
calculus. We also specify whether the calculus features \emph{cuts} (that is,
the combination of restriction and parallel composition corresponding with the
cut of linear logic) and continuations.
%
\CP~\cite{Wadler14,GayVasconcelos25} and \GV~\cite{GayVasconcelos10,Wadler14}
are well-known acronyms in the literature on session types. \SessionPI refers to
(variants of) the session-based $\pi$-calculus presented by
\citet{HondaVasconcelosKubo98} while \LinearPI refers to (variants of) the
linear $\pi$-calculus~\cite{KobayashiPierceTurner99}. Finally, \SCP is the
structural version of \CP introduced by \citet{SanoKavanaghPientka23} and
\Calculus is our calculus.
%
We emphasize that the actual calculus being formalised usually differs from
(typically, is a strict subset of) the one identified by the acronym and that
the same acronym may sometimes refer to different versions of the same calculus.
In particular, \GV in a logical setting is described by \citet{Wadler14} but its
first (non-logical) version is due to \citet{GayVasconcelos10}.

@@

The three middle columns of \Cref{tab:comparison} report the relevant
qualitative aspects of the formalisations, namely the management of typing
contexts, the representation of channels and whether processes are intrinsically
or extrinsically typed.
